An Improved Approximate-Bayesian Model-choice Method for Estimating Shared Evolutionary History

### 2014-02-21 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
More editing of dpp ms.

\begin{linenomath}
I implement two prior probability distributions over the space
of all possible discrete divergence models (\divTimeIndexVector).
The first simply gives all possible unordered partitions of \npairs{} elements
equal probability
\begin{equation}
    p(\divTimeIndexVector) = \frac{1}{\integerPartitionNum{\npairs{}}},
    \label{eq:divModelPriorUniform}
\end{equation}
i.e., a discrete uniform prior over all the integer partitions of \npairs{}
(discrete divergence models).
I denote this prior as
$\divTimeIndexVector \sim \priorUniform$.
\end{linenomath}

@@

\subsection{Differences between this model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsubsection{Ordering of taxon-specific summary statistics}
As alluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, the summary statistics are re-ordered by descending values of average
pairwise differences between the descendant populations
\citep[$\pi_b$;][]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of discrete
divergence models \divTimeIndexVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

\subsection{Assessing power}
I evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
\citet{Oaks2012}.
The only difference from the prior models used in the validation analyses above
is the prior on the concentration parameter \concentrationParam for the
\modelDPP model.
Rather than the prior
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(2,2)}$,
I use the prior
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(1.5,18.1)}$
over the discrete divergence models for the model \modelDPP.
The reason for this difference is because for all of the power simulations, I
used a simulated data structure identical to that of the empirical dataset of
Philippine vertebrates analyzed by \citet{Oaks2012}, which consists of 22 pairs
of populations.
For each of the four models, I generated $2\e6$ samples from the prior.
I then simulated datasets from three series of models
(Table~\ref{tabPowerModels}).
One series of models, which I denote \powerSeriesOld, is identically
distributed as \modelOld except that the divergence times for each of the 22
pairs of populations are randomly drawn from a series of uniform distributions,
$U(0, \divt{max})$, where \divt{max} was set to: 0.2, 0.4, 0.6, 0.8, 1.0, and
2.0, in \globalcoalunit generations.
A second series of models, \powerSeriesUniform, is identically distributed as
\modelUniform and \modelDPP except that the 22 divergence times are randomly
drawn from the same series of uniform priors as above.
The third series of models, \powerSeriesExp, is also identically distributed as
\modelUniform and \modelDPP except that the 22 divergence times are randomly
drawn from a series of of exponential distributions:
$Exp(mean=0.058)$, $Exp(mean=0.115)$, $Exp(mean=0.173)$, $Exp(mean=0.231)$,
$Exp(mean=0.289)$, and $Exp(mean=0.577)$.
These exponential distributions have the same variance as their uniform
counterparts in the first two series of models.

>>>>

\begin{linenomath}
I implement two prior probability distributions over the space
of all possible divergence models (\divTimeIndexVector).
The first simply gives all possible unordered partitions of \npairs{} elements
equal probability
\begin{equation}
    p(\divTimeIndexVector) = \frac{1}{\integerPartitionNum{\npairs{}}},
    \label{eq:divModelPriorUniform}
\end{equation}
i.e., a discrete uniform prior over all the integer partitions of \npairs{}
(unordered divergence models).
I denote this prior as
$\divTimeIndexVector \sim \priorUniform$.
\end{linenomath}

@@

\subsection{Differences between this model and the original \texttt{msBayes} model}
\subsubsection{The prior on divergence models}
One of the key differences between my model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
I denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, I place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
I introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides an investigator with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsubsection{Ordering of taxon-specific summary statistics}
As alluded to in the model description, \msb does not maintain the order of the
taxon-specific summary statistics \alignmentSS{}{} within each \ssVector{}.
Rather, the summary statistics are re-ordered by descending values of average
pairwise differences between the descendant populations
\citep[$\pi_b$;][]{NeiLi1979,Huang2011}.
This has the advantage of reducing the sample space of the number of
divergence models \divTimeIndexVector, but there are at least two disadvantages.
First, additional information in the data is lost.
By discarding the identity of the \npairs{} pairs of populations, all
pair-specific information about the amount of data (e.g., the number gene
copies collected from each of the populations [\popSampleSize{1}{} and
\popSampleSize{2}{}], the number of loci, and the length of the loci), and the
taxon- and locus-specific parameters (\hkyModel{}{},
\mutationRateScalarConstant{}{}, \ploidyScalar{}{}, and
\locusMutationRateScalar{}) is lost.
Second, the results are more difficult to interpret, because they can be no
longer be directly associated to the taxa under study.

@@

\subsection{Assessing power}
I evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
\citet{Oaks2012}.
For all power simulations, I used a data structure identical to that of the
empirical dataset of Philippine vertebrates analyzed by \citet{Oaks2012}, which
consists of 22 pairs of populations.
Due to the larger number of pairs, I used a different hyperprior on the
concentration parameter for the \modelDPP model; I used a prior of
$\divTimeIndexVector \sim \priorDPP{\sim Gamma(1.5,18.1)}$ over divergence
models for the model \modelDPP.
All other aspects of the four models in Table~\ref{tabPriors} were identical to
those used in the validation analyses described above.

For each of the four models, I generated $2\e6$ samples from the prior.
I then simulated datasets from three series of models
(Table~\ref{tabPowerModels}).
One series of models, which I denote \powerSeriesOld, is identically
distributed as \modelOld except that the divergence times for each of the 22
pairs of populations are randomly drawn from a series of uniform distributions,
$U(0, \divt{max})$, where \divt{max} was set to: 0.2, 0.4, 0.6, 0.8, 1.0, and
2.0, in \globalcoalunit generations.
A second series of models, \powerSeriesUniform, is identically distributed as
\modelUniform and \modelDPP except that the 22 divergence times are randomly
drawn from the same series of uniform priors as above.
The third series of models, \powerSeriesExp, is also identically distributed as
\modelUniform and \modelDPP except that the 22 divergence times are randomly
drawn from a series of of exponential distributions:
$Exp(mean=0.058)$, $Exp(mean=0.115)$, $Exp(mean=0.173)$, $Exp(mean=0.231)$,
$Exp(mean=0.289)$, and $Exp(mean=0.577)$.
These exponential distributions have the same variance as their uniform
counterparts in the first two series of models.
